\documentclass[11pt]{article}

\usepackage{geometry}
\geometry{margin=1in}

\usepackage{amsmath,amssymb}
\usepackage{hyperref}

\title{The Babylon Continuation: Deterministic Replay and Refusal-First Execution in Language-Driven Systems}
\author{J. M. Bookbinder\\
Witness Grade Analytics (WGA)\\
Atlanta, Georgia, USA}
\date{March 2026}

\begin{document}

\maketitle

\begin{abstract}
All claims in this paper are contingent on artifact verification. No claim is authoritative without a corresponding run bundle and decision receipt.

Modern systems increasingly rely on language as an execution surface for initiating, modifying, and validating process state. This paper introduces a deterministic continuation control framework for language-driven systems, enforcing a decision model with precedence HALT > INCOMPLETE > PASS.

The method requires replay-verifiable artifacts, cryptographic integrity checks, and explicit refusal records when continuation cannot be justified. Language is treated as telemetry rather than narrative, and execution is permitted only when structural admissibility is preserved under the declared evaluation scope, with deterministic replay required where verified.
\end{abstract}

\section{Introduction}

Language has become an operational interface for modern systems, including software pipelines, financial systems, and artificial intelligence workflows. However, current systems lack a deterministic method for verifying whether execution should continue once language has altered system state.

This paper presents a continuation control architecture in which language is treated as structured signal input, and system execution is governed by replay-verifiable evidence rather than narrative interpretation.

\section{Continuation Decision Model}

The system evaluates continuation under strict precedence:

\begin{center}
\textbf{HALT $>$ INCOMPLETE $>$ PASS}
\end{center}

\begin{itemize}
\item HALT: Structural violation or missing invariant prevents continuation.
\item INCOMPLETE: Required evidence or artifacts are missing.
\item PASS: All constraints satisfied within the declared evaluation scope.
\end{itemize}

\section{Deterministic Replay Requirement}

Admissible execution under replay requires reproducibility under identical inputs. This condition is enforced when replay certification is within the declared evaluation scope. The system enforces:

\begin{itemize}
\item Artifact hashing (SHA-256)
\item Run bundle integrity
\item Replay equivalence
\end{itemize}

Any mismatch between original and replayed execution results in immediate HALT.

\section{Refusal-First Execution Control}

If continuation cannot be justified, the system must emit a refusal record rather than proceed.

\begin{itemize}
\item No silent failure
\item No narrative override
\item Explicit obstruction documentation
\end{itemize}

Refusal is treated as a valid and required system outcome.

\section{Language as Telemetry}

Language is not treated as explanation but as a structured signal capable of altering system state. All language inputs must be:

\begin{itemize}
\item Captured as artifacts
\item Hash-bound
\item Replay-tested
\end{itemize}

This prevents semantic drift and ensures that meaning is tied to verifiable structure.

\section{Conclusion}

The Babylon Continuation framework establishes a deterministic, artifact-first method for governing execution in language-driven systems. By enforcing replay-verifiable evidence and refusal-first control, the system ensures that continuation is granted only when structurally admissible.

Deterministic replay, cross-host reproduction, and execution equivalence remain unverified conditions in the current execution instance.

\end{document}

\section{Bound Run Artifact (Sealed Instance)}

This document is bound to a specific run instance under the continuation-control system.

\subsection{Artifact Binding}

\begin{itemize}
\item Artifact: \texttt{paper.pdf}
\item Artifact SHA256: \texttt{c913b1bfb0f250782cbba3da158cef53f39c1cc316a4cef3e96ad638d32d4c0e}
\item Source: \texttt{paper.tex}
\item Source SHA256: \texttt{e1c4beb01ea9f009ae30a6dc41e01307898f1f581e240749d489f9c3a5483200}
\item Run Bundle: \texttt{run_bundle.json}
\item Run Bundle SHA256: \texttt{8144ba79122d28b4efb95d69ec205c9f1a1d5ef73c550b105c667fd7d132e9bc}
\item Decision Receipt SHA256: \texttt{87b60f9572c30bf546988d8ac375276ff16088219dc30f68e76cd54fbeeebdbe}
\end{itemize}

\subsection{Decision Scope}

This artifact is sealed under the following continuation conditions:

\begin{itemize}
\item Decision: PASS
\item Scope: artifact sealing
\item Claim Strength: artifact sealed, not replay certified
\end{itemize}

\subsection{Observer-Address Condition}

Admissibility is not global. This artifact is:

\begin{itemize}
\item admissible for sealing under local observer address
\item not yet admissible for cross-address replay
\end{itemize}

\begin{quote}
An artifact counts insofar as it supports a declared function at a declared observer address.
\end{quote}

\subsection{Open Conditions}

The following conditions remain unresolved:

\begin{itemize}
\item deterministic replay across hosts
\item execution equivalence under replay
\item cross-address admissibility
\end{itemize}

